---
title: "Supplementary Material: High-Throughput Bioimage Phenotyping and Computer Vision Workflows Using pliman"
subtitle: "A Python-Free R Framework Integrating Foundation Models, Few-Shot Learning, and Native ONNX GPU Acceleration"
author: "Tiago Olivoto et al."
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
    code-tools: true
    code-copy: true
    code-overflow: wrap
    highlight-style: github
---

# Overview and Architecture

This document provides a comprehensive, reproducible reference guide for the computer vision and deep learning modules in **pliman** (Plant Image Analysis in R). 

### Eliminating the Python Dependency Barrier
In conventional R workflows, deploying deep learning models (such as YOLO, SAM, or CLIP) requires maintaining Python virtual environments, Conda installations, and framework bindings (e.g., `reticulate` with PyTorch or TensorFlow). This dependency chain frequently leads to software fragility, CUDA driver mismatches, and reproducibility failures on non-specialist hardware.

**pliman** resolves this bottleneck by implementing a native **C++ ONNX Runtime engine** directly compiled into the R package. Key advantages include:
1. **Zero Python or Conda requirements:** Everything runs natively in R.
2. **Universal GPU Acceleration:** Hardware acceleration is provided through Microsoft DirectML and ONNX Runtime, delivering high-speed tensor execution across NVIDIA, AMD, and Intel GPUs on Windows and Linux, as well as multi-threaded CPU execution on macOS.
3. **End-to-End Pipeline:** Covers everything from zero-shot foundation models and 1-click few-shot interactive dataset annotation, to full YOLO training, model quantization, focus stacking, root system architecture phenotyping, and real-time video tracking.

```{r setup, message=FALSE, warning=FALSE}
library(pliman)

# Set random seed for reproducibility
set.seed(42)
```

---

# Foundation Models and Deep Learning Inference

Foundation models are large neural networks trained on broad visual data that can be applied directly to novel biological scenes without prior dataset collection or model retraining.

## 4x Deep Super-Resolution and Salient Instance Segmentation

When images contain low-resolution details (e.g., small foliar lesions or micro-structures), Real-ESRGAN super-resolution enhances edge clarity prior to instance segmentation.

```{r superres-instance-seg, eval=FALSE}
# 1. Import image and apply 4x Deep Learning Super-Resolution (Real-ESRGAN)
img_leaf <- image_import("leaf.jpg") |> 
  image_superres_dl(scale = 4, engine = "gpu")

# 2. Salient instance segmentation with rainbow color palette
seg_leaf <- image_segment_dl(
  img = img_leaf,
  model = "silueta",
  type = "highlight",
  bbox = TRUE,
  rainbow = TRUE,
  engine = "gpu"
)

# 3. Visualize feature activation and saliency heatmap
plot_yolo_heatmap(seg_leaf, col_palette = "viridis")
```

## Open-Vocabulary Zero-Shot Detection and Prompting (Grounded-SAM)

Researchers can segment specific biological organs, symptoms, or foreground targets using natural language prompts without any training data:

```{r grounded-sam, eval=FALSE}
img_canopy <- image_import("canopy.jpg")

# Segment specific targets via text prompts
canopy_seg <- image_segment_dl(
  img = img_canopy,
  model = "grounded-sam",
  prompt = "green leaf, necrotic lesion",
  type = "highlight",
  box_threshold = 0.25,
  text_threshold = 0.25,
  bbox = TRUE,
  rainbow = TRUE,
  engine = "gpu"
)

# Inspect detected bounding boxes and confidence scores
head(canopy_seg$boxes)
```

## Oriented Bounding Boxes (YOLO OBB)

In agricultural remote sensing, drone imagery, and forestry, objects often appear at arbitrary orientations. Oriented Bounding Boxes (OBB) provide tightly bounded rotated rectangles:

```{r yolo-obb, eval=FALSE}
img_aerial <- image_import("aerial_field.jpg")

# Detect rotated bounding boxes
obb_res <- image_detect_dl(
  img = img_aerial,
  model = "yolo26",
  task = "obb",
  conf_threshold = 0.50,
  engine = "gpu"
)

plot(obb_res)
```

## Cellular and Dense Nuclei Segmentation (StarDist)

For microscopy phenotyping, seed embryos, and dense cellular structures, **pliman** provides StarDist integration:

```{r stardist, eval=FALSE}
img_cells <- image_import("microscopy_nuclei.png")

cells_seg <- image_segment_dl(
  img = img_cells,
  model = "stardist",
  engine = "gpu"
)

plot(cells_seg)
```

## Salient Foreground Extraction (U2-Net)

Salient object extraction isolates foreground biological specimens (e.g., leaves, fruits, grains) from complex backgrounds:

```{r u2net-batch, eval=FALSE}
# Import a list of specimen images
specimens <- list(
  coffee1 = image_import("coffee_bean1.jpg"),
  coffee2 = image_import("coffee_bean2.jpg"),
  soybean = image_import("soybean_seed.jpg"),
  fruit   = image_import("citrus_fruit.jpg")
)

# Batch background removal
cleaned_specimens <- lapply(specimens, function(im) {
  image_segment_dl(im, model = "u2net", engine = "gpu", alpha = 0.8)
})

# Display mosaic of segmented foregrounds
image_combine(cleaned_specimens, ncol = 4, mar = 0)
```

---

# Few-Shot Learning and Automated Dataset Labeling

Manual bounding-box and polygon annotation is a major bottleneck in agricultural and veterinary machine learning. **pliman** implements an interactive human-in-the-loop few-shot pipeline using Meta's Segment Anything 2.1 (SAM 2.1 / PerSAM).

## Benchmark Dataset: OpenCows2020 (Open-Access Cattle Phenotyping)

To ensure full reproducibility without copyright restrictions, this case study utilizes the open-access **OpenCows2020** benchmark dataset (University of Bristol, licensed under CC-BY 4.0), comprising overhead imagery of Holstein dairy cattle in handling chutes:
- **Dataset Structure**: `OpenCows2020/img/` containing high-resolution top-down RGB frames.
- **Reference**: Andrew et al. (2020). Visual identification of individual Holstein-Friesian cattle via deep metric learning. *Computers and Electronics in Agriculture*.

## Interactive Few-Shot Prototype Fitting (`yolo_fewshot_fit`)

The researcher clicks on only 1–3 representative animal instances across one or more reference images. The engine extracts high-dimensional visual prototypes and automatically segments all matching objects across the scene.

```{r fewshot-fitting, eval=FALSE}
# Reference images representing visual cattle variation
ref_images <- c(
  "OpenCows2020/img/000010.jpg",
  "OpenCows2020/img/000020.jpg",
  "OpenCows2020/img/000030.jpg"
)

# Interactive model fitting
# 1. Click on 1-3 exemplar animals in the plot window and press <Esc>
# 2. SAM 2.1 segments instances and immediately displays the feedback overlay
# 3. Use the interactive console menu:
#    - [Enter]: Accept detections and proceed to the next image
#    - [r]: Remove false positives (click on masks/boxes to delete)
#    - [a]: Add missed objects (click on missed instances to incorporate them)
#    - [t <val>]: Fine-tune confidence threshold on the fly
#    - [s]: Resample points for this image
model_cows <- yolo_fewshot_fit(
  img = ref_images,
  label = "cow",
  task = "segment",        # "segment" for polygons or "detect" for bounding boxes
  method = "sam",
  conf_threshold = 0.55,
  rainbow = TRUE,
  engine = "gpu"
)

# Inspect the multi-image visual prototypes and segmentation overlay
plot(model_cows, rainbow = TRUE)
```

## Batch Dataset Auto-Labeling (`yolo_dataset_autolabel`)

Once the few-shot visual prototypes are calibrated, propagate annotations across hundreds of unannotated field images in seconds, generating standard YOLO dataset structures (`images/`, `labels/`, and `data.yaml`):

```{r batch-autolabel, eval=FALSE}
# Automatically annotate entire directory into YOLO format
ds_cows <- yolo_dataset_autolabel(
  input_dir  = "OpenCows2020/img",
  model      = model_cows,
  output_dir = "dataset_cows_yolo",
  max_images = 150,        # Annotate a representative subset or all frames
  train_prop = 0.80,       # 80% train, 20% validation
  engine     = "gpu",
  review     = TRUE,       # Exports visual review overlays
  rainbow    = TRUE
)

# Inject negative/background images to prevent false positives
yolo_dataset_add_background(
  dir    = ds_cows,
  source = "backgrounds/",
  n      = 20
)

# Apply data augmentation (flips, rotations, lighting adjustments)
yolo_dataset_augment(ds_cows)

# Preview generated dataset annotations
yolo_dataset_preview("dataset_cows_yolo", n = 4)
```

## Traditional and Assisted Contour Export (`yolo_dataset_export`)

For high-contrast images (e.g., leaves on a white background, seeds on a scanner tray), classical morphological algorithms can directly generate YOLO-format polygonal datasets with Bézier/contour smoothing:

```{r dataset-export, eval=FALSE}
# Generate YOLO segmentation dataset using morphological contours
dataset_vitis <- yolo_dataset_export(
  img         = "vitis_leaves/",
  dir         = "dataset_vitis",
  task        = "segment",
  class_names = "grape_leaf",
  train_prop  = 0.80,          # 80% train, 20% validation
  n_points    = 150,           # Polygonal vertex count per contour
  index       = "GRAY",
  watershed   = FALSE,
  opening     = 5,
  return_exact = TRUE
)

# Preview dataset
yolo_dataset_preview("dataset_vitis", n = 4)
```

---

# Custom YOLO Model Training and Validation

Researchers can train cutting-edge YOLO architectures directly inside R without switching to external Python environments.

## Object Detection and Instance Segmentation Training

```{r yolo-training, eval=FALSE}
# Train a custom YOLO26 instance segmentation model
model_trained <- yolo_train(
  data        = "dataset_cows_yolo/data.yaml",
  model       = "yolo26n-seg",
  output_name = "cows_seg.onnx",
  output_dir  = "models_cows",
  epochs      = 50,
  batch       = 8,
  imgsz       = 640
)

# Plot training loss curves, Precision-Recall curves, and confusion matrix
plot(model_trained)
plot(model_trained, which = "pr_curve")
plot(model_trained, which = "confusion_matrix")
```

## Model Validation and Real-World Inference

Deploy the trained ONNX model directly within R:

```{r yolo-inference, eval=FALSE}
img_test <- image_import("OpenCows2020/img/000041.jpg")

# Run inference with the custom-trained model
res_test <- image_segment_dl(
  img            = img_test,
  model          = "models_cows/cows_seg.onnx",
  conf_threshold = 0.40,
  rainbow        = TRUE
)

plot(res_test)
```

## Deep Image Classification Pipeline

**pliman** also provides a streamlined workflow for image-level classification (e.g., disease diagnosis, botanical variety identification):

```{r classification-pipeline, eval=FALSE}
# 1. Structure raw image folders into train (70%), val (20%), and test (10%) splits
ds_flowers <- yolo_dataset_classify(
  src_dir = "flower_photos",
  out_dir = "dataset_flowers_yolo",
  split   = c(train = 0.70, val = 0.20, test = 0.10),
  seed    = 42
)

# Inspect dataset balance and summary
yolo_dataset_balance(ds_flowers)
plot(ds_flowers)

# 2. Train a classification model
model_flowers <- yolo_train(
  data   = ds_flowers,
  model  = "yolo11l-cls",
  epochs = 50,
  batch  = 16,
  imgsz  = 224
)

# 3. Classify a novel specimen
new_img <- image_import("flower_photos/sunflowers/sample1.jpg")
pred_class <- image_classify_dl(new_img, model = "models/flowers_cls.onnx")
plot(pred_class)
```

---

# Model Optimization for Edge Computing (INT8 Quantization)

For deployment on edge devices (e.g., agricultural robots, smart tractors, battery-powered camera traps), 8-bit integer quantization reduces model size by ~75% and significantly accelerates inference.

```{r int8-quantization, eval=FALSE}
# Quantize standard FP32 ONNX model to INT8
onnx_quantize(
  model_path  = "models_logs/logs_seg.onnx",
  output_path = "models_logs/logs_seg_int8.onnx"
)

# Benchmark execution latency: FP32 vs. INT8
test_sample <- image_import("toras/toras19.jpg")

latency_benchmark <- bench::mark(
  FP32 = image_detect_dl(test_sample, model = "models_logs/logs_seg.onnx", plot = FALSE, engine = "gpu"),
  INT8 = image_detect_dl(test_sample, model = "models_logs/logs_seg_int8.onnx", plot = FALSE, engine = "gpu"),
  iterations = 20,
  check = FALSE
)

summary(latency_benchmark)
```

---

# Real-Time Video Monitoring and Tracking

Video analysis enables dynamic monitoring of conveyor belts, livestock, and real-time field operations.

## Live Camera and Human Pose Estimation

Detect and track anatomical skeletons in real-time from webcam feeds:

```{r video-pose, eval=FALSE}
# Real-time pose estimation from connected webcam (device 0)
live_pose <- video_detect_dl(
  video          = 0,
  model          = "yolo26n-pose",
  engine         = "gpu",
  conf_threshold = 0.50,
  skeleton       = TRUE,      # Draws 17-point anatomical skeleton
  bbox           = TRUE,
  return_data    = TRUE
)
```

## Video Tracking and Trajectory Tracing

Track moving objects across video files with persistent identity assignment:

```{r video-tracking, eval=FALSE}
# Multi-object tracking on video file
tracked_video <- video_detect_dl(
  video       = "cows.mp4",
  model       = "yolo26x-seg",
  engine      = "gpu",
  fps         = 24,
  track       = TRUE,         # Enables persistent object ID tracking
  trail       = TRUE,         # Draws movement trajectory trails
  mask        = TRUE,
  rainbow     = TRUE,
  output      = "cows_tracked.mp4",
  return_data = TRUE
)

print(tracked_video)
plot(tracked_video)
```

## Conveyor Belt Threshold Counting

Automate the enumeration of seeds, fruits, or parts passing along a conveyor system:

```{r video-conveyor, eval=FALSE}
# Automated object counting crossing a virtual line
conveyor_data <- video_detect_dl(
  video            = "conveyor_system.mp4",
  model            = "yolo26n",
  engine           = "gpu",
  conf_threshold   = 0.70,
  roi                    = c(0.2, 0.8),                 # Monitored relative band [min, max]
  roi_orientation        = "vertical",                  # "vertical" or "horizontal"
  roi_mode               = "crop",
  count_line             = 0.5,                         # Normalized position (0 to 1)
  count_line_orientation = "vertical",                  # "vertical" or "horizontal"
  count_line_label       = "Belt Total",
  track                  = TRUE,
  output           = "conveyor_annotated.mp4",
  return_data      = TRUE
)

# Inspect enumerated count total
conveyor_data$counts
```

## Video Frame Extraction

```{r video-split, eval=FALSE}
# Extract frames from video at a constant frequency for dataset creation
video_split(
  video = "field_recording.mp4",
  dir   = "extracted_frames/",
  fps   = 2                  # Extract 2 frames per second
)
```

---

# Specialized Computational Imaging and Phenotyping

## Deep Learning Focus Stacking (IFCNN)

In macro photography and stereo-microscopy, limited depth of field causes foreground and background structures to blur. Deep convolutional focus stacking fuses multi-focal image series into an all-in-focus composite:

```{r focus-stacking, eval=FALSE}
# Load focus bracketed series
focus_series <- list(
  image_import("focus/plane_top.jpg"),
  image_import("focus/plane_middle.jpg"),
  image_import("focus/plane_bottom.jpg")
)

# Deep neural feature fusion via IFCNN
fused_composite <- process_focus_stack(
  images        = focus_series,
  fusion_method = "ifcnn",
  strategy      = "pyramid",
  engine        = "gpu"
)

plot(fused_composite)
```

## Panoramic Aerial Image Stitching

Stitch multi-angle aerial or ground photograph sequences into a panoramic orthomosaic:

```{r image-stitching, eval=FALSE}
# Import overlapping aerial photographs
drone_tiles <- image_import(pattern = "DJI_")

# Automatically align, warp, and blend tiles
field_mosaic <- image_stitch(
  drone_tiles,
  direction = "auto",
  blend     = TRUE
)

plot(field_mosaic)
```

## Root System Architecture (RSA) Phenotyping

Comprehensive architectural quantification of root systems:

```{r root-phenotyping, eval=FALSE}
img_root <- image_import("root_scan.png") |> 
  image_vertical()

# Comprehensive root morphological extraction
root_traits <- root_analyze(img_root)

# Multifaceted architectural trait visualizations
plot(root_traits, type = "hierarchy",   nodes = FALSE)
plot(root_traits, type = "skeleton",    nodes = FALSE, lwd = 2)
plot(root_traits, type = "overlay",     cex_nodes = 0.9, nodes = FALSE)
plot(root_traits, type = "classes")
plot(root_traits, type = "depth")
plot(root_traits, type = "angles")
plot(root_traits, type = "profile",     asp = 1)        # Spatial envelope and centroid
plot(root_traits, type = "topology")                    # Gravitropism: angle vs. depth
plot(root_traits, type = "diameter_dist")               # Diameter distribution and KDE
plot(root_traits, type = "orders")                      # Biomass distribution by root order
```

## Generative Image Synthesis (Latent Diffusion)

Generate synthetic biological training samples or background variations via diffusion models:

```{r diffusion-synthesis, eval=FALSE}
# Generate high-resolution synthetic botanical imagery
synthetic_field <- image_generate_dl(
  prompt   = "a mature wheat field under natural morning light, high resolution botanical photography",
  aspect   = "landscape",
  n_steps  = 4,
  superres = TRUE,
  engine   = "gpu"
)

plot(synthetic_field)
```

---

# Classical Morphology vs. Deep Learning Benchmark

To assist researchers in selecting the appropriate methodology, **pliman** allows side-by-side comparison between classical color-index thresholding and deep learning segmentation:

```{r classical-comparison, eval=FALSE}
img_sample <- image_import("standard_sample.jpg")

# 1. Classical color index segmentation (Otsu threshold on R channel)
classical_res <- analyze_objects(
  img_sample,
  index = "R",
  show_segmentation = TRUE
)

# Inspect classical shape statistics (area, perimeter, circularity)
head(classical_res$results)

# 2. Deep learning instance segmentation (SAM 2.1)
dl_res <- image_segment_dl(
  img_sample,
  model = "silueta",
  type = "highlight",
  rainbow = TRUE
)

# Comparison plot
image_combine(
  list(Classical = classical_res$plot, DeepLearning = dl_res),
  ncol = 2
)
```

---

# Session Information

```{r session-info}
sessionInfo()
```
